---
title: "Gapminder Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    social: [ "twitter", "facebook", "menu"]
    source_code: embed
---

```{r setup, include=FALSE}
library(flexdashboard)
library(knitr)
library(DT)
library(rpivotTable)
library(ggplot2)
library(plotly)
library(dplyr)
library(openintro)
library(highcharter)
library(ggvis)
library(readxl)
library(kableExtra)
```

Data
=============================================
<br>
<font size = "5">Gapminder Data -- Head & Tail</font>
<br>
```{r}
gapminder <- read.csv("gapminder.csv")

gapminder[c(1:6,1699:1704),1:6] %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = T) %>%
  pack_rows("Head", 1, 6, label_row_css = "background-color: #bebebe;") %>%
  pack_rows("Tail", 7, 12, label_row_css = "background-color: #bebebe;")
```
<br>
<br>
<font size = "5">Gapminder Summary</font>
<br>
```{r}
options(knitr.kable.NA = '')
summary(gapminder) %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = T)
```


Boxplots
=============================================

### **GDP Per Capita** by *Continent & Country (1952-2007)*

```{r}
plot_ly(gapminder, x = ~gdpPercap, y = ~country) %>%
add_boxplot(color = ~continent) %>%
layout(yaxis = list(title = ""),
xaxis = list(title = "GDP Per Capita"))
```


Scatterplot 1 {data-navmenu="Scatterplots"}
=============================================


### **GDP Per Capita** by *Life Expectancy (1952-2007)*

```{r}
p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) + geom_point() + facet_wrap(~year)
ggplotly(p)
```

Scatterplot 2 {data-navmenu="Scatterplots"}
=============================================

### **Logarithmic GDP Per Capita** by *Life Expectancy (1952-2007)*

```{r}
p <- ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop)) + geom_point() + facet_wrap(~year) + scale_x_log10()
ggplotly(p)
```

Tree-Map
=============================================

### **Average Life Expectancy** by *Country*

```{r}
tree_data <- gapminder %>%
  group_by(country) %>%
  summarise(avg_lifeExp = mean(lifeExp)) %>%
  arrange(-avg_lifeExp) %>%
  mutate(parents = "Average Life Expectancy 1952-2007") %>%
  ungroup() 
  
  plot_ly(data = tree_data,
          type = "treemap",
          values = ~avg_lifeExp,
          labels = ~country,
          parents =  ~parents,
          domain = list(column=0),
          name = "Avg Life Expectancy 1952-2007",
          textinfo = "label")
```

Sources
=============================================

<font size ="6">**Data Sources**</font>

<font size ="5">
Data used for the creation of this dashboard includes data compiled from various sources by Gapminder Foundation. Datasets can be located at [gapminder.org](https://gapminder.org).
</font>
<br>

<font size ="6">**References**</font>

<font size ="5">
Gapminder (2021a). GDP per capita, constant PPP dollars: v26 [Data set]. Gapminder Foundation. http://www.gapminder.org/data/

Gapminder (2021b). GM-Population - Dataset - v7 [Data set]. Gapminder Foundation. http://www.gapminder.org/data/

Gapminder (2021c). _GM-Life Expectancy- Dataset - v12 [Data set]. Gapminder Foundation. http://www.gapminder.org/data/
</font>


