---
title: "Programming Tools & Techniques in Data Management"
author: "A. Malinish"
date: "2020-11-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required Packages

```{r}
library(psych)
```


## Normal Distribution Density Plot

In this section, a standard normal distribution is generated and displayed as a density plot. 

```{r}
x1 = seq(from = -6, to = 6, by = 0.01)
y1 = dnorm(x1)
plot(x1, y1, col = 7)
```

## Boxplot

This R code chunk first examines the structure of the *"airquality"* dataset and converts its **Month** variable from an integer to factor variable. A boxplot that plots **Month** against **Temp** is then generated. The boxplot reveals that **Month** = 7 has the highest mean **Temp**; meanwhile, **Month** = 9 appears to have the largest range of **Temp** values. 

```{r}
data(airquality)
str(airquality)
airquality$Month <- as.factor(airquality$Month)
attach(airquality)
boxplot(Temp ~ Month, col = "skyblue")
```

## Creating and Storing Functions

A new function is created and stored as **FunData** in this step. **FunData** is used to create and specify values for new variables and return the variables as a dataframe. **FunData** is then used to created the dataframe, *df*.

```{r}
FunData <- function(){
  ID = 1:20
  Gender = rep(c("M","F"),10)
  Age = c(19,20,19,20,21,22,23,19,18,24,21,22,19,20,21,23,21,22,20,18)
  Weight = rnorm(20, mean = 130, sd = 5)
  Height = rnorm(20, mean = 70, sd = 4)
  return(data.frame(ID,Gender,Age,Weight,Height))
}

df <- FunData()
df
```

## Computing Mean Values

Mean values for **Age**, **Weight**, and **Height** by **Gender** from the dataframe *df* are calculated in this step using the tapply() function. Mean values are as follows:
<br>

* **Age**, **Gender**
  + Female: 21.0
  + Male: 20.2

* **Weight**, **Gender**
  + Female: 131.8713
  + Male: 128.8837
  
* **Height**, **Gender**,
  + Female: 71.20746
  + Male: 71.07279
  
```{r}
attach(df)
tapply(Age, Gender, mean)

tapply(Weight, Gender, mean)

tapply(Height, Gender, mean)
```

## Creating Calculated Variable

The dataset, *bodyfat* (Penrose, 1985), was imported into R and stored as *bodyfat* for this section. The structure of *bodyfat* (Penrose, 1985) includes 252 observations comprised of 15 variables. An additional variable, **bmi**, was created by performing manual calculations for **bmi** and storing them as a vector. The cbind() function was then used to add the vector of **bmi**'s to *bodyfat*. A summary of the variables shows that **bmi** has a mean value of 25.94.

```{r}
bodyfat <- read.csv("bodyfat.csv")
str(bodyfat)

bmi <- 703*bodyfat$weight/bodyfat$height^2
bodyfat <- cbind(bodyfat,bmi)

head(bodyfat)
summary(bodyfat)
```

## Correlation Matrix

The pairs.panels() function from the "psych" function was used to generate a correlation matrix for all of the variables in the *bodyfat* dataset. 

The correlation matrix of **density**, **bodyfat**, **weight**, **height**, and **bmi** indicates that the strongest correlation exists between **density** and **bodyfat** (Pearson correlation = -0.99). This Pearson correlation value represents a negative correlation, meaning that as one variable increases in value, the other variable decreases in value. Meanwhile, the strongest positive correlation exists between **bodyfat** and **weight** (Pearson correlation = 0.61). This type of correlation suggests that both variables increase and decrease with one another.

```{r}
pairs.panels(bodyfat[,c(1,2,4,5,16)])
```
<br>

## References

From "Generalized Body Composition Prediction Equation for Men Using Simple Measurement Techniques [Data set]," by Penrose, K., Nelson, A. and Fisher, A. (1985). *Medicine and Science in Sports and Exercise, 17*(189). http://dx.doi.org/10.1249/00005768-198504000-00037

Revelle, W. (2020). psych: Procedures for Psychological, Psychometric, and Personality Research. *Northwestern University*. https://CRAN.R-project.org/package=psych